Lossy compression of DNA sequencing quality data

### 2015-05-01 manuscript/q-compression.tex
Cleaned up text a bit.  Added more to discussion.

Despite the fact that polynomial functions have significantly lower
number of parameters (i.e. one to six coefficients) than a read-length
string of raw quality values, the necessity of using floating point
numbers to store coefficients greatly limits the compression potential
of the method. In order to get meaningful compression on
single-precision four-byte floating point numbers, one would have to
relax on the least-squares approximation constraint to obtain
compressible values on the byte level which is outside the scope of
this study.

@@

\textsc{QualComp} is a lossy compression tool that models quality
values as a multivariate Gaussian distribution, computing the mean and
covariance for each position in the read\citep{Ochoa:2013rt}. Once the
model parameters are calculated, they are stored by the decoder to
later reconstruct a \emph{representative} quality value.
\textsc{QualComp} takes as input a user-specified rate (bits per read)
and then poses an optimization problem of how to allot these bits for
a given position while minimizing the \textsc{mse}. The quality values
can be clustered before hand to produce more accurate models.

@@

Certain short read alignment tools use the quality value information
when finding potential alignments. Bowtie2 (version 2.2.3) was used to
evaluate the different decompressed \textsc{fastq} files. Bowtie2 uses
quality values written in the \textsc{fastq} files when mapping reads
against a reference genome. The original uncompressed and decompressed
\textsc{fastq} files were mapped with Bowtie2 against
\textit{Rhodobacter sphaeroides} reference genome. The generated
\textsc{sam} file for each compressing approach were compared with the
uncompressed \textsc{sam} file. The total, shared and unique
proportional numbers of mapped reads are calculated with respect to
the uncompressed \textsc{sam} matches as shown in table
\ref{tab:aligner}. Additionally, to monitor the effect of quality
values on mapping in general, Bowtie2 was adjusted so that the maximum
and minimum mismatch penalty were equivalent to maximum and minimum
quality scores (with parameters: --mp 6,6 and --mp 2,2 respectively).

@@

@@

We have simply provided here the initial steps in analyzing the effect
of lossy compression on quality values using high-coverage bacterial,
human, and mouse data sets. A more direct comparison against
additional lossy compression tools, such as
\textsc{SlimGene}~\citep{Kozanitis:2011kl}, needs to be
performed. Furthermore, other types of sequencing data can be analyzed
apart from the Illumina data examined here. For example, the PacBio
sequencing instruments produce very long reads (with average read
lengths on the order of 10 kbp), but with the trade-off of having a
high error-rate ($\sim$15\%). Unlike the class of quality values we
have examined here, the distribution of erroneous bases is relatively
uniform~\cite{Ferrarini:2013vf}. The assembly complexity of bacterial
genomes can be greatly simplified, producing near complete genome
assemblies, by utilizing a single run of these long
reads~\citep{Koren:2013ye}. If long read sequencing technologies such
as PacBio become more widely adopted, it would be of huge benefit to
examine the potential of lossy compression algorithms on not only the
quality values, but the biological sequencing data themselves.

@@

>>>>

Despite the fact that polynomial functions have significantly lower
number of parameters (i.e. one to six coefficients in our approach) than a read-length
string of raw quality values, the necessity of using floating point
numbers to store coefficients greatly limits the compression potential
of the method. In order to get meaningful compression on
single-precision four-byte floating point numbers, one would have to
relax on the least-squares approximation constraint to obtain
compressible values on the byte level which is outside the scope of
this study.

@@

\textsc{QualComp} is a lossy compression tool that models quality
values as a multivariate Gaussian distribution, computing the mean and
covariance for each position in the read\citep{Ochoa:2013rt}. Once the
model parameters are calculated, they are stored by the decoder to
later reconstruct a \emph{representative} quality value.
\textsc{QualComp} takes as input a user-specified rate (bits per read)
and then poses an optimization problem of how to allot these bits for
a given position within the read while minimizing the \textsc{mse}. The quality values
can be clustered before hand to produce more accurate models.

@@

Certain short read alignment tools use the quality value information
when finding potential alignments. Bowtie2 (version 2.2.3) was used to
evaluate the different decompressed \textsc{fastq} files. Bowtie2 uses
quality values written in the \textsc{fastq} files when mapping reads
against a reference genome. The original uncompressed and decompressed
\textsc{fastq} files were mapped with Bowtie2 against
\textit{Rhodobacter sphaeroides} reference genome. The generated
\textsc{sam} file for each compressing approach were compared with the
uncompressed \textsc{sam} file. The total, shared and unique
proportional numbers of mapped reads are calculated with respect to
the uncompressed \textsc{sam} matches as shown in Table
\ref{tab:aligner}. Additionally, to monitor the effect of quality
values on mapping in general, Bowtie2 was adjusted so that the maximum
and minimum mismatch penalty were equivalent to maximum and minimum
quality scores (with parameters: --mp 6,6 and --mp 2,2 respectively).

@@

It is important to highlight that the downstream applications dictate which lossy compression methods are acceptable.
As highlighted earlier, 2-bin encoding is able to retain information about low quality bases that occur within stretches of high quality bases, whereas the polynomial regression methods tend to smooth these qualities.
In addition to aligning less reads that the 2-bin encoding, the polynomial regression methods may also be hindering downstream tools that use the alignment results.
A \textsc{snp} detection pipeline will commonly use the quality values to record if a potential difference from the reference is the result of genetic variation or sequencing error.
Erroneously marking a base as higher quality, may lead to the base being incorrectly labeled as a \textsc{snp}.
2-bin encoding is acceptable for alignment; however, it performs quite poorly in terms of assembly.
In the case of assembly, the 7th-degree polynomial regression method outperforms 2-bin encoding in nearly all metrics.

@@

We have simply provided here the initial steps in analyzing the effect
of lossy compression on quality values using high-coverage bacterial,
human, and mouse data sets. A more direct comparison against
additional lossy compression tools, such as
\textsc{SlimGene}~\citep{Kozanitis:2011kl}, needs to be
performed. Furthermore, other types of sequencing data can be analyzed
apart from the Illumina data examined here. For example, the PacBio
sequencing instruments produce very long reads (with average read
lengths on the order of 10 kbp), but with the trade-off of having a
high error-rate ($\sim$15\%). Unlike the class of quality values we
have examined here, the distribution of erroneous bases is relatively
uniform~\citep{Ferrarini:2013vf}. The assembly complexity of bacterial
genomes can be greatly simplified, producing near complete genome
assemblies, by utilizing a single run of these long
reads~\citep{Koren:2013ye}. If long read sequencing technologies such
as PacBio become more widely adopted, it would be of huge benefit to
examine the potential of lossy compression algorithms on not only the
quality values, but the biological sequencing data themselves.

@@

\section{Availability}
Implementations of the compression methods written in Python and R and pipeline to reproduce the results are available at \url{https://github.com/cmhill/q-compression}
